An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines

### 2020-08-22 [2019-11]_HPC_(quintic_spline)/paper/paper.tex
Deleted unnecessary files, renamed some things, cleaning up mostly.

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
\begin{align*}
h_i &= k_{i+1} - k_{i}, \quad
u(t) = 3t^2 - 2t^3, \quad
p(t) = t^3 - t^2,\\
\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
p\big((x-k_i)/h_i\big).
\end{align*}
Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$. However, such a spline has too many \textit{wiggles} for most applications. \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$, otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$. Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

@@

\begin{table}
  \centering
  \begin{tabular}{c|c c c|c c c|c c c}
    \hline
    \multirow{2}{*}{$n$}
      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
      & min & median & max & min & median & max & min & median & max \\
    \hline
    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
    \hline
  \end{tabular}
  \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data. Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
  \label{table:e3_results}
\end{table}

@@

\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has
interests in numerical analysis, mathematical programming,
bioinformatics, and data science.  He has been involved with the
organization of HPCS since 2000.

\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
studying computer science under Dr. Layne Watson.

\textbf{\uppercase{Li Xu}} is a Ph.D. candidate at VPI\&SU studying
statistics under Dr. Yili Hong.

\textbf{\uppercase{Yueyao Wang}} is a Ph.D. candidate at VPI\&SU
studying statistics under Dr. Yili Hong.

\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) is an
associate professor of Statistics at VPI\&SU.

>>>>

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
\begin{equation}\begin{aligned}
h_i &= k_{i+1} - k_{i}, \quad
u(t) = 3t^2 - 2t^3, \quad
p(t) = t^3 - t^2,\\
\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
p\big((x-k_i)/h_i\big).
\end{aligned}\end{equation}
Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$. However, such a spline has too many \textit{wiggles} for most applications. \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$, otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$. Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

@@

\begin{table}
  \centering
  \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data. Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
  \label{table:e3_results}
  \begin{tabular}{c|c c c|c c c|c c c}
    \hline
    \multirow{2}{*}{$n$}
      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
      & min & median & max & min & median & max & min & median & max \\
    \hline
    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
    \hline
  \end{tabular}
\end{table}

@@

\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has
interests in numerical analysis, mathematical programming, and data
science.  He has been involved with the organization of HPCS since
2000. \email{ltw@vt.edu}

\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
studying under Dr. Layne Watson. \email{thchang@vt.edu}

\textbf{\uppercase{Li Xu}} is a Ph.D. candidate at VPI\&SU studying
statistics under Dr. Yili Hong. \email{lix1992@vt.edu}

\textbf{\uppercase{Yueyao Wang}} is a Ph.D. candidate at VPI\&SU
studying under Dr. Yili Hong. \email{yueyao94@vt.edu}

\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) is an
associate professor of Statistics at VPI\&SU. \email{yilihong@vt.edu}
